---
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ldsR)
library(testthat)
library(fs)
```

# Introduction

ldsR exposes five functions:

1\. `celltype_analysis()` to estimate partitioned heritability for many
annotations one at a time, while adjusting for a common set of
annotations

2\. `ldsc_h2()` to estimate heritability for a single trait

3\. `ldsc_rg()` to estimate the genetic correlation between two traits

4\. `munge()` to apply quality control filters to a summary statistics
file

6\. `partitioned_h2()` to estimate partitioned heritability - OBS,
correct enrichemnt estimates for overlapping annotations has not yet
been implemented.

# Use ldsR to directly estimate heritability inside R

Estimate heritability with one line inside R.

`ldsc_h2()` has only one required argument, a dataframe with the columns
`SNP`, `Z` and `N`. The `ref-ld-chr` and `w-ld-chr` files are shipped
with the ldsR package, but you can also provide your own. By default
ldsR uses the ldscores derived from the European subset of 1000 Genomes.

```{r}
# example file, 100 000 rows to be lightweight.
df <- arrow::read_parquet(system.file("extdata", "sumstats.parquet", package = "ldsR"))


tdf <- dplyr::rename(df, Z = Z.x, N = N.x)
ldsc_h2(tdf)
```

# Genetic correlations

To calculate the genetic correlation between two traits, pass both
data.frames to `ldsc_rg()`. Note that there are now five mandatory
columns in each data.frame:

`SNP` `Z` `N` `A1` `A2`.

`A1` and `A2` are required to align the direction of the `Z` value
across two summary statistics.

```{r}
sumstat1 <- dplyr::rename(df, Z = Z.x, N = N.x) |> 
  dplyr::mutate(A1 = "A", A2 = "G")
sumstat2 <- dplyr::rename(df, Z = Z.y, N = N.y) |> 
  dplyr::mutate(A1 = "A", A2 = "G")

ldsc_rg(sumstat1, sumstat2)

```

`ldsc_rg()` provides an interface to easily run make genetic
correlations with an index trait.

for `sumstats2` you can provide a list of data.frames, and ldsc_rg will
calculate the genetic correlation with each summary statistic in the
list. If you provide a named list, the names will be saved in the
`trait2` column.

```{r}
sumstats <- list(
  "trait1" = sumstat2,
  "trait2" = sumstat2,
  "trait4" = sumstat2
)
          
ldsc_rg(sumstat1, sumstats)
        
```

# Munge

The `munge()` function can be used to mimic some of the filters of the
munge_sumstats.py file from ldsc.

```{r}
munge(sumstat1)
```

# The updated ldscore format

ldsR uses an updated format for ldscores, to reduce size and file
complexity. To run [partitioned_h2()] or [celltype_analysis()], you
provide a filepath to a directory containing two files: `ld.parquet` and
`annot.parquet`.

```{r}
fs::dir_tree(system.file("extdata", "baseline1.1_test", package = "ldsR"))

```

## ld.parquet
`ld.parquet` contains the LD scores for each annotation
```{r}
arrow::read_parquet(system.file("extdata", "baseline1.1_test/ld.parquet", package = "ldsR")) |> dplyr::slice_head(n = 5)
```

## annot.parquet
`annot.parquet` contains the annotations and M50 and M values.
```{r}
arrow::read_parquet(system.file("extdata", "baseline1.1_test/annot.parquet", package = "ldsR")) |> dplyr::slice_head(n = 5) 
```

# Partitioned heritability

To estimate the partitioned heritability, `partitioned_h2()` is
available, which requires the `ldscore_dir` argument, a directory
containing the partitioned ldscore files `ld.parquet` and
`annot.parquet` which has the M50 and M values. OBS, for overlapping
annotations, the functionality to adjust the enrichment estimate has not
yet been implemented. Therefore the enrichment estimate should not be
used.

```{r}
partitioned_h2(
  sumstat1,
  ldscore_dir = system.file("extdata", "baseline1.1_test", package = "ldsR")
)
```

# Cell-type analysis

Another common usage is a version of partitioned heritability sometimes
referred to as "cell-type analysis".

For [celltype_analysis()] one extra argument is required:
`covariate_dir`. `covariate_dir` will be used to read in a set of
LDscores that will be used as covariates when estimating heritability
enrichment. The ldscores in `ldscores_dir` will be used to estimate the
enrichment for each annotation, one at a time, while adjusting for *all*
annotations in `covariate_dir`.

```{r}
celltype_analysis(
  sumstat1,
  covariate_dir = system.file("extdata", "baseline1.1_test", package = "ldsR"),
  ldscore_dir = system.file("extdata", "superclusters", package = "ldsR")
)

```
