---
 ? ---
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ldsR)
library(testthat)
library(fs)
library(dplyr)
```

# Introduction

ldsR is a R rewrite of the [ldsc](https://github.com/bulik/ldsc) software. The aim is to provide a significantly simplified user experience.

Currently, ldsR is able to reproduce all features of ldsc, except:

1\. LDscore calculation

2\. Quantitative annotations in partitioned heritability.

The showcase here is based on a small example file of 100,000 rows, with effect sizes from schizophrenia and bipolar disorder.

```{r}
df <- arrow::read_parquet(system.file("extdata", "sumstats.parquet", package = "ldsR"))
head(df)
```

# Munge

To perfectly reproduce LDSC results, the input summary statistics needs to filtered in an identical manner. `munge()` mirrors the functionality of `munge_sumstats.py`. Munge requires the following columns:

`SNP` `A1` `A2` `Z` `N`

```{r}
# create alleles so that we can showcase how munge works
test_df <- mutate(df, A1 = "G", A2 = "T", Z = Z.x, N=N.x)
test_df <- munge(test_df)
```

# SNP heritability

`ldsc_h2()` estimates the SNP-heritability for a single trait, one at a time. By default, regression weights and LDscores are from the European 1000G reference panel.

The only required argument is the summary statistics. If you want to use other regression and LDscore weights (for example from another ancestry, you can use the `weights` argument and the `M` argument).

```{r}
ldsc_h2(test_df)
```

For binary traits, it's often useful to convert the heritability estimate to the liability scale. This can be done using the `pop_prev` and `sample_prev` arguments.

In this particular case, the N column in the example file reflects the total number of cases and controls. In such a case, we need both the population prevalence and sample prevalence

```{r}
n_cont <- 77344
n_cas <- 53300
s_prev <- n_cas/(n_cas + n_cont)
ldsc_h2(test_df, pop_prev = 0.01, sample_prev = s_prev)


```

Ideally, for estimation of heritability, the per SNP sample size should be calculated using the effective sample size summed across cohorts, see [here](https://pmc.ncbi.nlm.nih.gov/articles/PMC10066905/). If the N column corresponds to the effective sample size, only the `pop_prev` argument is required (a 50% sample prevalence is assumed).

```{r}
eff_n <- calc_effective_n(n_cas, n_cont)
test_df$N <- eff_n
ldsc_h2(test_df, pop_prev = 0.01, sample_prev = s_prev)
```

Lastly, you can estimate the observed scale heritability first, and then calculate the liability scale heritability at your convenience using `liability_h2()`.

```{r}
liability_h2(0.37, 0.01, 0.4)
```

# Genetic correlations

ldsR also implements the estimation of genetic correlations in `ldsc_rg()`. Regression weights and LDscores are from the European 1000G reference panel, same as with `ldsc_h2()`.

Technically, `ldsc_h2()` only requires SNP, Z and N.

`ldsc_rg()` on the other hand requires both `A1` and `A2`, to be able to align Z scores to the same reference allele.

Using the example data:

```{r}
sumstat1 <- dplyr::rename(df, Z = Z.x, N = N.x) |> 
  dplyr::mutate(A1 = "A", A2 = "G")
sumstat2 <- dplyr::rename(df, Z = Z.y, N = N.y) |> 
  dplyr::mutate(A1 = "A", A2 = "G")

ldsc_rg(sumstat1, sumstat2)

```

Additionally, `ldsc_rg()` attemps to make it easier to calculate genetic correlations among many traits.

sumstats2 can be both a `dplyr::tibble()`, and a list containing multiple `dplyr::tibble()`'s. If a list is passed to the `sumstats2` argument, the genetic correlation will be estimated between `sumstats1` and all entries in the list.

```{r}
sumstats <- list(
  "trait1" = sumstat2,
  "trait2" = sumstat2,
  "trait4" = sumstat2
)
          
ldsc_rg(sumstat1, sumstats)
        
```

If the list is not named, the names will correspond to the index in the list.

```{r}
sumstats <- list(
  sumstat2,
  sumstat2,
  sumstat2
)
          
ldsc_rg(sumstat1, sumstats)

```

# Partitioning heritability

`ldsR` deviates from `ldsc` in the file format used for the partitioned LD scores.

The example files used below are subset of the full LDscores, which are available [here](https://zenodo.org/records/15194124).

The original baseline annotations in ldsc LDscore format are available [here](https://zenodo.org/records/10515792).

The primary difference between ldsR and ldsc partitioned ldscores is the number of files. LDSC splits out the files across chromosomes and data-type, resulting in 88 files per data set (22 chromosomes, .ldscore, .annot, .m50, .m5_50).

ldsR collapses this to three files, stored in parquet files.

```{r}
dir_tree(system.file("extdata", "superclusters/", package = "ldsR"))

```

`annot.parquet` contains information on the number of SNPs, ie the M50 and M5_50 files per annotation

```{r}
arrow::read_parquet(system.file("extdata", "superclusters/annot.parquet", package = "ldsR")) |>
  head()

```

`ld.parquet` contains the LDscores per annotation

```{r}
arrow::read_parquet(system.file("extdata", "superclusters/ld.parquet", package = "ldsR")) |> 
  head()
```

`annot_ref.parquet` contains the SNP to annotation mapping for each annotation.

A value of 1 means that the SNP is mapped to that annotation.

```{r}
arrow::read_parquet(system.file("extdata", "superclusters/annot_ref.parquet", package = "ldsR")) |> 
  head()
```

The main function to partition heritability is `partition_h2()`.

1.  `sumstats` is the first argument, corresponding to a dataframe with SNP, Z and N. Use `munge()` first.
2.  `ldscore_dirs` takes a vector of filepaths, to directories with partitioned ldscores
3.  `subset_annots` allows for flexible subsetting of annotations to use in the model. Default is NULL; which means that all of the annotions will be used in the model
4.  `overlapping_annotations` which will result in the enrichment for each annotation also being estimated.

In this first example we estimate the partitioned heritability across the 4 annotations present in this example data.

```{r}
# use the dataframe we munged earlier

partition_h2(
  test_df,
  ldscore_dirs = system.file("extdata", "baseline1.1_test/", package = "ldsR"),
)

```

A common analysis is to estimate the enrichment of a particular annotation

Perhaps we are interested in doing "cell-type analysis", testing the enrichment of a specific annotation while adjusting for the 53 baseline annotations. With `partition_h2()`, we simply add the filepath to the directory with additional ldscores.

Here we use a subsetted example of LDscores derived from human brain cell-types.

To be more specific in which LDscores we want to include in the model, we need pass a list of annotation names that will be included in the model. For this we need to know what the annotation names are per ldscore dataset. Here we use `get_annot_names()`, which will simply return the names of all annotations in a ldscore dataset.

```{r}
all_ldscores <- c(
  system.file("extdata", "baseline1.1_test/", package = "ldsR"), 
  system.file("extdata", "superclusters/", package = "ldsR")
  )

baseline_names <- get_annot_names(all_ldscores[1])
supercluster_names <- get_annot_names(all_ldscores[2])

```

```{r}
baseline_names
```

```{r}
supercluster_names
```

With this, we specify which annotations to include in the model through `subset_annots`.

```{r}
partition_h2(
  test_df,
  ldscore_dirs = all_ldscores,
  # here we test the association with hippocampal neurons
  subset_annot = c(baseline_names,"amygdala_excitatory")
) |> 
  filter(annot == "amygdala_excitatory")
```

With this setup we can fit any model we might like, for example by adding additional cell-types to run conditional analysis.

```{r}
partition_h2(
  test_df,
  ldscore_dirs = all_ldscores,
  # here we test the association with hippocampal neurons
  subset_annot = c(baseline_names,"amygdala_excitatory","astrocyte")
) |> 
  filter(annot %in% c("amygdala_excitatory","astrocyte"))
```

Lastly, we can use `overlapping_annotations=TRUE` to estimate the enrichment of each annotation.

```{r, eval = FALSE}
partition_h2(
  test_df,
  ldscore_dirs = all_ldscores,
  subset_annot = c(baseline_names,"amygdala_excitatory","astrocyte"),
  overlapping_annotations=TRUE
) |> 
  filter(annot %in% c("amygdala_excitatory","astrocyte"))

```

# Cell-type analysis

Lastly, `celltype_analysis()` is a convenience function for estimating the association across several annotations, while adjusting for a set of "baseline" annotations.

1.  `sumstat`
2.  `covariate_dir` filepath to annotations that are used as controls
3.  `ldscore_dir` filepath to annotations that are to be tested, one at a time while adjusting for all annotations in covariate_dir

`celltype_analysis()` will loop over each annotation in ldscore_dir, testing it's association while adjusting for all annotations in covariate_dir.

```{r}
celltype_analysis(
  sumstat = test_df,
  covariate_dir = all_ldscores[1],
  ldscore_dir = all_ldscores[2]
  
)
```

# 
